A comparison of single-cell trajectory inference methods: towards more accurate and robust tools

### 2018-09-18 manuscript/paper.Rmd
update paper

In this study, we evaluated a total of →`r n_methods_evaluated`← methods in terms of cellular →ordering, topology, scalability, and user friendliness←. Our results indicate that existing methods are typically better at ordering the cells rather than inferring the correct trajectory topology, and that there is no "one-fits-all" method. Instead, the choice of method depends on the trajectory topology present in the data. →To aid users with selecting the most appropriate method, we therefore developed a set of guidelines, available as an interactive app at `r print_url("http://35.239.15.98")`,← While the field certainly has matured significantly since its inception, further progress is still necessary in order to cope with increasingly complex and novel use cases. We hope to spearhead such developments and the field as a whole, by making our evaluation pipeline easily extensible and freely available at `r print_url("https://www.github.com/dynverse/dynverse")`.

@@

<!-- Inclusion criterion -->
Methods were excluded from the evaluation based on several criteria: `r glue::glue_collapse(glue::glue("({non_inclusion_reasons$footnote}) {non_inclusion_reasons$long}"), ", ")`. The discussions on individual methods can be found at `r print_url("https://github.com/dynverse/dynmethods/issues?q=label:unwrappable")` This resulted in the inclusion of  `r n_methods_evaluated` methods in the evaluation.

<!-- Components -->
The selected methods were dissected into a set of algorithmic components. Each component can have a significant impact on the performance, scalability, and output. Across all methods, these components can be broadly grouped into two stages; (i) conversion to a simplified representation using dimensionality reduction, clustering or graph building and (ii) ordering the cells along the simplified representation [@cannoodt_computational_2016]. Components are frequently shared between different algorithms. For example, minimal spanning trees (MST), used in the first single-cell RNA-seq trajectory inference methods [@trapnell_dynamics_2014], is shared by almost half of the methods we evaluated.

→
### Method wrappers
Each method was wrapped within a docker and singularity container (available at `r print_url("https://github.com/dynverse/dynverse/tree/master/methods")`). These containers are automatically built on both Singularity Hub (`r print_url("https://singularity-hub.org/")`) and Docker Hub (`r print_url("https://hub.docker.com/u/dynverse")`).
For each method, we wrote a wrapper script based on example scripts or tutorials provided by the authors (as mentioned in the respective wrapper scripts). This script will read in the input data (as described below), run the method, and output the files required to construct a trajectory (as described below). We also created a script to generate an example dataset, which is used to automatically test every method using travis continuous integration (`r print_url("https://travis-ci.org/dynverse")`).

We used the github issues system to contact the authors of each method, and asked for feedback on the wrappers, the metadata, and the quality control. About ⅓ of authors responded, and we improved the wrappers based on their feedback. These discussion can be viewed on github: `r print_url('https://github.com/dynverse/dynmethods/issues?q=label:"method+discussion"')`. 
←
### Method input
As input, we provided for each method either the raw count data (after cell and gene filtering) or normalised expression values, based on the description in the methods documentation or from the study describing the method.
<!-- Prior information -->
A large portion of the methods requires some form of prior information (e.g. a start cell) in order to be executable. Other methods optionally allow to exploit certain prior information, though this is not required. Prior information can be supplied as a starting cell from which the trajectory will originate, a set of important marker genes, or even a grouping of cells into cell states. Providing prior information to a TI method can be both a blessing and a curse. In one way, prior information can help the method to find the correct trajectory among many, equally likely, alternatives. On the other hand, incorrect or noisy prior information can bias the trajectory towards current knowledge. Moreover, prior information is not always easily available, and its subjectivity can therefore lead to multiple equally plausible solutions, restricting the applicability of such TI methods to well studied systems.

@@

* **Start cells** The identity of one or more start cells. For both real and synthetic data, a cell was chosen which was the closest (in geodesic distance) to each milestone with only outgoing edges. For ties, one random cell was chosen. For cyclic datasets, a random cell was chosen.
* **End cells** The identity of one or more end cells. Similar as the start cells, but now for every state with only ingoing edges.
* **# end states** Number of terminal states. Number of milestones with only ingoing edges.
* **Grouping** For each cell a label to which state/cluster/branch it belongs. For real data, the states from the gold/silver standard. For synthetic data, each milestone was seen as one group, and cells were assigned to their closest milestone.
* **# branches** Number of branches/intermediate states. For real data, the number of states in the gold/silver standard. For synthetic data, the number of milestones.
→
* **Discrete time course** For each cell a time point from which it was sampled. If available, directly extracted from the gold standard, otherwise the geodesic distance from the root milestone was used. For synthetic data, four discrete timepoints were chosen, at which the cells were “sampled” to provide a time course information reflecting the one provided in real experiments.
* **Continuous time course** For each cell a time point from which it was sampled. For real data this was equal to the discrete time course, for synthetic data we used the internal simulation time of each simulator.
←
### Common trajectory model
Due to the absence of a common format for trajectory models, most methods return a unique set of output formats with few overlaps. →We therefore post-processed the output of each method into a common probabilistic trajectory model← (`r ref("sfig", "overview_wrapping", "a")`). This model consisted of three parts: (i) The milestone network represents the overall network topology, and contains edges between different milestones and the length of the edge between them. (ii) The milestone percentages contain, for each cell, its position between milestones, and sums for each cell to one. (iii) The trajectory model also allows for regions of delayed commitment, where cells are allowed to be positioned between three or more connected milestones. Regions must be explicitly defined in the trajectory model. Per region, one milestone must be directly connected to all other milestones →of the region.←

@@

More information on how each method was wrapped can be found within the comments of each wrapper script.
←

→
### Control methods
We added several control positive and negative control methods: **random**, which generates a random network with 50 milestones according to a Barabási–Albert model, **identity**, which returns the gold standard, and **shuffle**, which will place the cells randomly on the gold standard milestone network.
### Off-the-shelf methods
For baseline performance, we added several "off-the-shelf" TI methods which can be run using just a couple of commands in R.←

@@

## Real datasets
We gathered `r sum(list_datasets()$source == "real")` real datasets by searching for "single-cell" at the Gene Expression Omnibus and selecting those datasets in which the cells are sampled from different stages in a dynamic process (`r ref("stable", "datasets")`). The scripts to download and process these datasets are available on our repository ([https://github.com/dynverse/dynbenchmark/tree/master/scripts/01-datasets](https://github.com/dynverse/dynbenchmark/tree/master/scripts/01-datasets)). Whenever possible, we preferred to start from the raw counts data. These raw counts were all normalised and filtered using a common pipeline, discussed later.

→For each dataset, we extracted a reference trajectory, consisting of two parts: the cellular grouping (milestones) and the connections between these groups (milestone network). The cellular grouping was provided by the authors of the original study, and we classified it as a gold standard when it was created independently from the expression matrix (such as from cell sorting, the origin of the sample or the time it was sampled) or silver standard otherwise (usually by clustering the expression values). To connect these cell groups, we used the original study to determine the network which the authors proposed or otherwise found to be the most likely. In the end, each group of cells was placed on a milestone, having percentage = 1 for that particular milestone. The known connections between these groups were used to construct the milestone network. If there was time data available, we used this as the length of the edge, otherwise we set all the lengths equal to one. ←

@@

The workflow to generate synthetic data is based on the well established workflow used in the evaluation of network inference methods [@schaffter_genenetweaversilicobenchmark_2011; @marbach_wisdom_2012] and consists of four main steps: network generation, simulation, gold standard extraction and simulation of the scRNA-seq experiment. At every step, we tried to mirror real regulatory networks, while keeping the model simple and easily extendable. For every synthetic dataset, we used a random real dataset as a reference dataset (from those described earlier), making sure the number of variable genes and cells were similar.
#### Network generation

@@

Because each cellular simulation follows the trajectory at its own speed, knowing the exact position of a cell within the trajectory topology is not straightforward. Furthermore, the speed at  which simulated cells make a decision between two or more alternative paths is highly variable. To estimate a cell's position during a simulation within the trajectory topology, we therefore used the known progression of the modules, as a backbone. We smoothed the expression in each simulation using a rolling mean with a window of 50 time steps, and then calculated the average module expression along the simulation. We used dynamic time warping, implemented in the dtw R package [@giorgino_computingvisualizingdynamic_2009; @tormene_matchingincompletetime_2009], with an open end to align a simulation to all possible module progressions, and then picked the alignment which minimised the normalised distance between the simulation and the backbone. In case of cyclical trajectory topologies, the number of possible milestones a backbone could progress through was limited to 20.

@@

Each quality aspect received a weight depending on how frequently it was found in several papers and online sources which discuss tool quality (`r ref("stable", "qc_checks")`). This was to make sure that more important aspects, such as the open source availability of the method, outweighed other aspects, such as the availability of a graphical user interface. Within each aspect, we also assigned a weight to the individual questions being investigated (`r ref("stable", "qc_checks")`). For calculating the final score, we weighed each of the six categories equally.
→
## Scalability
←
## Trajectory types
We classified all possible trajectory topologies into distinct trajectory types, based on topological criteria (`r ref("fig", "ti_evaluation_overview", "c")`). These trajectory types start from the most general trajectory type, a disconnected directed graph, and move down (within a directed acyclic graph structure), progressively becoming more simple until the two basic types: linear and cyclical. A disconnected directed graph is a graph in which only one edge can exist between two nodes. A directed graph is a disconnected graph in which all nodes are (indirectly) connected. A directed acyclic graph is a graph containing no cycles. A rooted tree is a directed acyclic graph containing no convergences (no nodes with in-degree higher than 1). A rooted binary tree is a rooted tree with every node having maximally two outgoing edges. A convergence is a directed acyclic graph in which only one node has a degree larger than one and this same node has an indegree of one. A multifurcation is a rooted tree in which only one node has a degree larger than one. A bifurcation is a multifurcation in which only one node has a degree equal to 3. A directed linear graph is a graph in which no node has a degree larger than 3. Finally, a directed cycle is a directed graph in which no node has a degree larger than 3 but contains one cycle. For every directed trajectory type, a corresponding undirected trajectory type can also be defined, with directed acyclic graphs and directed graphs merging to undirected graph, and convergence and bifurcation merging to a simple fork. In most cases, a method which was able to detect a complex trajectory type, was also able to detect less complex trajectory types, with the only exception being DPT (limited to bifurcating trajectories).

>>>>

In this study, we evaluated a total of →`r n_methods_evaluated`← methods in terms of cellular →ordering, topology, scalability, and user friendliness←. Our results indicate that existing methods are typically better at ordering the cells rather than inferring the correct trajectory topology, and that there is no "one-fits-all" method. Instead, the choice of method →mostly← depends on the trajectory topology present in the data. →To aid users with selecting the most appropriate method, we therefore developed a set of guidelines, available as an interactive app at `r print_url("http://35.239.15.98")`.← While the field certainly has matured significantly since its inception, further progress is still necessary in order to cope with increasingly complex and novel use cases. We hope to spearhead such developments and the field as a whole, by making our evaluation pipeline easily extensible and freely available at `r print_url("https://www.github.com/dynverse/dynverse")`.

@@

<!-- Inclusion criterion -->
Methods were excluded from the evaluation based on several criteria: `r glue::glue_collapse(glue::glue("({non_inclusion_reasons$footnote}) {non_inclusion_reasons$long}"), ", ")`. →The discussions on individual methods can be found at `r print_url("https://github.com/dynverse/dynmethods/issues?q=label:unwrappable")`. In the end, we included  `r n_tools` tools in the evaluation, constituting `r n_methods_evaluated` methods.←

<!-- Components -->
The selected methods were dissected into a set of algorithmic components. Each component can have a significant impact on the performance, scalability, and output. Across all methods, these components can be broadly grouped into two stages; (i) conversion to a simplified representation using dimensionality reduction, clustering or graph building and (ii) ordering the cells along the simplified representation [@cannoodt_computational_2016]. Components are frequently shared between different algorithms. For example, minimal spanning trees (MST), used in the first single-cell RNA-seq trajectory inference methods [@trapnell_dynamics_2014], is shared by almost half of the methods we evaluated.
→
### Method wrappers
Each method was wrapped within a docker and singularity container (available at `r print_url("https://github.com/dynverse/dynverse/tree/master/methods")`). These containers are automatically built on both Singularity Hub (`r print_url("https://singularity-hub.org/")`) and Docker Hub (`r print_url("https://hub.docker.com/u/dynverse")`).
For each method, we wrote a wrapper script based on example scripts or tutorials provided by the authors (as mentioned in the respective wrapper scripts). This script reads in the input data (as described below), runs the method, and outputs the files required to construct a trajectory (as described below). We also created a script to generate an example dataset, which is used to automatically test every method using travis continuous integration (`r print_url("https://travis-ci.org/dynverse")`).

We used the github issues system to contact the authors of each method, and asked for feedback on the wrappers, the metadata, and the quality control. About ⅓ of authors responded, and we improved the wrappers based on their feedback. These discussions can be viewed on github: `r print_url('https://github.com/dynverse/dynmethods/issues?q=label:"method+discussion"')`. 
←
### Method input
As input, we provided each method with either the raw count data (after cell and gene filtering) or normalised expression values, based on the description in the method documentation or from the study describing the method.
<!-- Prior information -->
A large portion of the methods requires some form of prior information (e.g. a start cell) in order to be executable. Other methods optionally allow to exploit certain prior information. Prior information can be supplied as a starting cell from which the trajectory will originate, a set of important marker genes, or even a grouping of cells into cell states. Providing prior information to a TI method can be both a blessing and a curse. In one way, prior information can help the method to find the correct trajectory among many, equally likely, alternatives. On the other hand, incorrect or noisy prior information can bias the trajectory towards current knowledge. Moreover, prior information is not always easily available, and its subjectivity can therefore lead to multiple equally plausible solutions, restricting the applicability of such TI methods to well studied systems.

@@

* **Start cells** The identity of one or more start cells. For both real and synthetic data, a cell was chosen which was the closest (in geodesic distance) to each milestone with only outgoing edges. For ties, one random cell was chosen. For cyclic datasets, a random cell was chosen.
* **End cells** The identity of one or more end cells. Similar as the start cells, but now for every state with only ingoing edges.
* **# end states** Number of terminal states. Number of milestones with only ingoing edges.
* **Grouping** For each cell a label to which state/cluster/branch it belongs. For real data, the states from the gold/silver standard. For synthetic data, each milestone was seen as one group, and cells were assigned to their closest milestone.
* **# branches** Number of branches/intermediate states. For real data, the number of states in the gold/silver standard. For synthetic data, the number of milestones.
→
* **Discrete time course** For each cell a time point from which it was sampled. If available, directly extracted from the gold standard, otherwise the geodesic distance from the root milestone was used. For synthetic data, four discrete timepoints were chosen, at which the cells were “sampled” to provide a time course information reflecting the one provided in real experiments.
* **Continuous time course** For each cell a time point from which it was sampled. For real data this was equal to the discrete time course, for synthetic data we used the internal simulation time of each simulator.
←
### Common trajectory model
Due to the absence of a common format for trajectory models, most methods return a unique set of output formats with few overlaps. →We therefore post-processed the output of each method into a common probabilistic trajectory model← (`r ref("sfig", "overview_wrapping", "a")`). This model consisted of three parts: (i) The milestone network represents the overall network topology, and contains edges between different milestones and the length of the edge between them. (ii) The milestone percentages contains, for each cell, its position between milestones, and sums for each cell to one. (iii) →The regions of delayed commitment, which are connection between three or more milestones. These must be explicitly defined in the trajectory model and per region, one milestone must be directly connected to all other milestones of the region.←

@@

More information on how each method was wrapped can be found within the comments of each wrapper script.
←
→
### Control methods
We added several control positive and negative control methods: **random**, which generates a random network with 50 milestones according to a Barabási–Albert model, **identity**, which returns the gold standard, and **shuffle**, which will place the cells randomly on the gold standard milestone network.
### Off-the-shelf methods
For baseline performance, we added several "off-the-shelf" TI methods which can be run using just a couple of commands in R.←

@@

## Real datasets
We gathered `r sum(list_datasets()$source == "real")` real datasets by searching for "single-cell" at the Gene Expression Omnibus and selecting those datasets in which the cells are sampled from different stages in a dynamic process (`r ref("stable", "datasets")`). The scripts to download and process these datasets are available on our repository (`r print_url("https://github.com/dynverse/dynbenchmark/tree/master/scripts/01-datasets")`). Whenever possible, we preferred to start from the raw counts data. These raw counts were all normalised and filtered using a common pipeline, discussed later.

→For each dataset, we extracted a reference trajectory, consisting of two parts: the cellular grouping (milestones) and the connections between these groups (milestone network). The cellular grouping was provided by the authors of the original study, and we classified it as a gold standard when it was created independently from the expression matrix (such as from cell sorting, the origin of the sample or the time it was sampled) or silver standard otherwise (usually by clustering the expression values). To connect these cell groups, we used the original study to determine the network which the authors validated or otherwise found to be the most likely. In the end, each group of cells was placed on a milestone, having percentage = 1 for that particular milestone. The known connections between these groups were used to construct the milestone network. If there was time data available, we used this as the length of the edge, otherwise we set all the lengths equal to one. ←

@@

The workflow to generate synthetic data is based on the well established workflow used in the evaluation of network inference methods [@schaffter_genenetweaversilicobenchmark_2011; @marbach_wisdom_2012] and consists of four main steps: network generation, simulation, gold standard extraction and simulation of the scRNA-seq experiment. At every step, we tried to mirror real regulatory networks, while keeping the model simple and easily extendable. →We simulated a total of ←
#### Network generation

@@

Because each cellular simulation follows the trajectory at its own speed, knowing the exact position of a cell within the trajectory topology is not straightforward. Furthermore, the speed at  which simulated cells make a decision between two or more alternative paths is highly variable. →We therefore first constructed a backbone expression profile for each branch within the trajectory. To do this, we first defined in which order the expression of the modules is expected to change, and then generated a backbone expression profile in which the expression of these modules increases and decreases smoothly between 0 and 1.  We also smoothed the expression in each simulation using a rolling mean with a window of 50 time steps, and then calculated the average module expression along the simulation. ← We used dynamic time warping, implemented in the dtw R package [@giorgino_computingvisualizingdynamic_2009; @tormene_matchingincompletetime_2009], with an open end to align a simulation to all possible module progressions, and then picked the alignment which minimised the normalised distance between the simulation and the backbone. In case of cyclical trajectory topologies, the number of possible milestones a backbone could progress through was limited to 20.

@@

Each quality aspect received a weight depending on how frequently it was found in several papers and online sources which discuss tool quality (`r ref("stable", "qc_checks")`). This was to make sure that more important aspects, such as the open source availability of the method, outweighed other aspects, such as the availability of a graphical user interface. Within each aspect, we also assigned a weight to the individual questions being investigated (`r ref("stable", "qc_checks")`). For calculating the final score, we weighed each of the six categories equally.
→
## Scalability
←
## Trajectory types
We classified all possible trajectory topologies into distinct trajectory types, based on topological criteria (`r ref("fig", "ti_evaluation_overview", "c")`). These trajectory types start from the most general trajectory type, a disconnected directed graph, and move down (within a directed acyclic graph structure), progressively becoming more simple until the two basic types: linear and cyclical. A disconnected directed graph is a graph in which only one edge can exist between two nodes. A directed graph is a disconnected graph in which all nodes are (indirectly) connected. A directed acyclic graph is a graph containing no cycles. A rooted tree is a directed acyclic graph containing no convergences (no nodes with in-degree higher than 1). A rooted binary tree is a rooted tree with every node having maximally two outgoing edges. A convergence is a directed acyclic graph in which only one node has a degree larger than one and this same node has an indegree of one. A multifurcation is a rooted tree in which only one node has a degree larger than one. A bifurcation is a multifurcation in which only one node has a degree equal to 3. A directed linear graph is a graph in which no node has a degree larger than 3. Finally, a directed cycle is a directed graph in which no node has a degree larger than 3 but contains one cycle. For every directed trajectory type, a corresponding undirected trajectory type can also be defined, with directed acyclic graphs and directed graphs merging to undirected graph, and convergence and bifurcation merging to a simple fork. In most cases, a method which was able to detect a complex trajectory type, was also able to detect less complex trajectory types, with →the exceptions being being DPT (limited to bifurcating trajectories), and STEMNET and FateID (unable to return linear trajectories).←
